---
title: "Data Visualisation with ggplot2 - Project"
output:
  html_document:
    df_print: paged
---

# AfterWork Data Science: Data Visualisation with R Project

## 1. Defining the Question

### a) Specifying the Data Analysis Question
Specify the reasearch question that you'll be answering. i.e. Provide strategy recommendations that will lead to revenue growth.

### b) Defining the Metric for Success
The solutions to the following questions will help us answer our research question:

* When is the best time of year to book a hotel room?
* When is the optimal length of stay in order to get the best daily rate?
* How will you know if a hotel was likely to receive a disproportionately high number of special requests?

### c) Understanding the context
Provide some background information....

### d) Recording the Experimental Design
Describe the steps/approach that you will use to answer the given question.

### e) Data Relevance
How relevant was the provided data?


## 2. Reading the Data

```{r}
# Load the data below
# Install and load tidyverse library
#install.packages("tidyverse")
library("tidyverse") 
# --- 
# Dataset url = 
# --- 
hotel_df <- read.csv("data/hotel_bookings.csv")

# 
```

```{r}
# Checking the first 5 rows of data
# ---
head(hotel_df, 5)
```

```{r}
# Checking the last 5 rows of data
# ---
tail(hotel_df,5)
```

```{r}
# Sample 10 rows of data
# --- 
sample(hotel_df, 10)
```

```{r}
# Checking number of rows and columns
# ---
dim(hotel_df)  
```

```{r}
# Checking datatypes
# ---
str(hotel_df) 
```

```{r}

```

Record your general observations below:

We have over 100,000 records and 32 columns.
All data are as of 2015


## 3. External Data Source Validation
The data is originally from the article Hotel Booking Demand Datasets, by Nuno Antonio, Ana Almeida, and Luis Nunes for Data in Brief, Volume 22, February 2019.


## 4. Data Preparation
### Performing Data Cleaning
```{r}
# Checking datatypes and missing entries of all the variables
# ---
is.na(hotel_df)
str(hotel_df)

```

We observe the following from our dataset:

Children column has some records with missing values
There are 3 datatypes, int, char and num in our dataset

```{r}
# Checking how many duplicate rows are there in the data
# ---
hotel_df[duplicated(hotel_df), ]
```

We choose to keep the duplicates because we don't have a unique identifier to actually proof that we have duplicates.

```{r}
# Checking if any of the columns are all null
# ---
colSums(is.na(hotel_df))
```

We observe the following from our dataset:

Children column has some null records

```{r}
# Checking if any of the rows are all null
# ---
is.na(hotel_df) 
```

We observe the following from our dataset:

We do not have a lot of missing data or rows with nulls, therefore we can use the data.

```{r}
# Checking the correlation of the features through the use of 
# visualizations the correlation using heatmap
# ---
sample(hotel_df)
```

We observe the following from our dataset:

Observation 1
Observation 2

```{r}
# Dropping company column because it has alot of missing values 
# and we won't need to answer any of our questions
# ---
hotel_df = select(hotel_df, -c("company"))
names(hotel_df)

```

From the data variable description we see that the Distribution Channel categoricy that tells us about Booking distribution.

The term “TA” means “Travel Agents”
The term “TO” means “Tour Operators”
This allows us to fill the missing values in the agents column with TO

```{r}
# We replace the mising values i.e. for TO with
# ---
hotel_df$agent[is.na(hotel_df$agent)] <- 9
```

```{r}
# We drop rows where there is no adult, baby and child as 
# these records won't help us.
# ---
# means that adults is 0 or childe
hotel_df <- hotel_df[!(hotel_df$adults == 0)]
```

```{r}
# We replace missing children values with rounded mean value
# ---
# Hint i.e. use round()
# ---
hotel_df$children[is.na(hotel_df$children)] <- round(mean(hotel_df$children, na.rm = TRUE),0)
```

```{r}
# Checking for missing values in the dataframe
# ---
colSums(is.na(hotel_df))
```

```{r}
# Converting the datatypes of the following columns from float to integer
# i.e. children, company, agent
# ---
hotel_df$children <- as.integer(hotel_df$children)
hotel_df$agent <- as.integer(hotel_df$agent)
str(hotel_df)
```


## 5. Solution Implementation

### 5.a) Questions

```{r}
# 1. How many bookings were cancelled?
# ---
# Visualisation: Barplot
#Summary table

hotel_df %>%
  filter(is_canceled == 1) %>%
  group_by(arrival_date_month) %>%
  summarise(n = n())

# Plotting the chart
ggplot(hotel_df, aes(x = arrival_date_month, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Total cancellations per month", 
    x = "Month", 
    y = "Total Cancellations"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


```{r}
# 2. What was the booking ratio between resort hotel and city hotel?
# ---
# Barplot of booking ratio between resort hotel and city hotel

# Summary table
hotel_df1 <- hotel_df
hotel_df1 %>%
  count(hotel) %>%
  mutate(perc = n / nrow(hotel_df1)) -> hotel_df1
ggplot(hotel_df1, aes(x = hotel, y = perc*100)) + geom_bar(stat = "identity")
```


```{r}
# 3. What was the percentage of booking for each year?
# ---
# Summary table
hotel_df2 <- hotel_df
hotel_df2 %>%
  count(arrival_date_year) %>%
  mutate(perc = n / nrow(hotel_df2)) -> hotel_df2
ggplot(hotel_df2, aes(x = arrival_date_year, y = perc*100)) + geom_bar(stat = "identity")
```

```{r}
# 4. Which were the most busiest months for hotels?
# ---
# 
hotel_df3 <- hotel_df
hotel_df3 %>% 
  filter(is_canceled==0) %>%
  group_by(arrival_date_month) %>%
  summarize(n = n()) %>%
  arrange(desc(n))
# Plotting the chart
ggplot(hotel_df3, aes(x = arrival_date_month, y = is_canceled )) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Busiest Months", 
    x = "Month", 
    y = "Total Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```


```{r}
# 5. From which top 3 countries did most guests come from?
# --- 
hotel_df4 <- hotel_df
hotel_df4 <- hotel_df4 %>% 
  filter(is_canceled==0) %>%
  group_by(country) %>%
  summarize(n = n()) %>%
  arrange(desc(n))

```

```{r}
# 6.a) How long do most people stay in hotels?
#   b) By city and resort? Separate the data by hotel
# ---
hotel_df_1 <- hotel_df
hotel_df_2 <- hotel_df

# By City Hotel Weeknights
hotel_df_1 %>% 
  filter(hotel == 'City Hotel') %>%
  group_by(stays_in_week_nights) %>%
  summarise(n = n()) %>%
  arrange(desc(n))

# By City Hotel Weekend nights
hotel_df_1 %>% 
  filter(hotel == 'City Hotel') %>%
  group_by(stays_in_weekend_nights) %>%
  summarise(n = n()) %>%
  arrange(desc(n))

# By Resort Hotel Weeknights
hotel_df_1 %>% 
  filter(hotel == 'Resort Hotel') %>%
  group_by(stays_in_week_nights) %>%
  summarise(n = n()) %>%
  arrange(desc(n))

# By Resort Hotel Weekend nights
hotel_df_1 %>% 
  filter(hotel == 'Resort Hotel') %>%
  group_by(stays_in_weekend_nights) %>%
  summarise(n = n()) %>%
  arrange(desc(n))



```


```{r}
# 7. Which was the most booked accommodation type (Single, Couple, Family)?
# ---
# Assumptions:
# Single (1 adult with no children or babies)
# Couple (2 adults with no children or babies)
# Family (at least 1 adult with either babies or children)
#
#Single
single <- hotel_df %>% 
  filter(adults ==1 & children == 0 & babies == 0) %>%
  summarise(n = n())
#Couple
couple <- hotel_df %>%
  filter(adults == 2 & children == 0 & babies == 0) %>%
  summarise(n = n())
#Family
family <- hotel_df %>%
  filter(adults >= 1 & (children >= 1 | babies >= 1)) %>%
  summarise(n = n())

accomdation <- c("Single", "Couple", "Family")
bookings <- c(single$n, couple$n, family$n)
n_hotel_df <- data.frame(accomdation,bookings)
n_hotel_df

ggplot(data=n_hotel_df, aes(x="", y=bookings, fill=accomdation)) +
  geom_bar(
    stat="identity", width=0.5, color = "white"
  ) +
  coord_polar("y") + 
  geom_text(
    aes(label = paste0(round(bookings / sum(bookings) * 100, 1), "%")), position = position_stack(vjust = 0.5)
  ) +
  scale_fill_brewer(palette="Set1") +   
  labs(
    fill = "Accomodation", 
    title = "Accommodation Types"
  ) +
  theme_void()
```
Highest accommodation type from the analysis is couple at 72% followed by Single at 19.9% and finally Family at 8%.

### 5.b) Recommendations
From the above analysis, below are our recommendations:
Try and increase family offers in order to attract more bookings. We can also try and add more activities(e.g. rave parties and excursions) with the Single package that will attract more singles to book rooms.
Seems a lot of people did not stay in during the weekend nights, we can try and add more entertainment joints like clubs and bars to encourage more people to stay in and enjoy in the hotel. This will increase our revenue streams.


## 6. Challenging your Solution
In this step, we review our solution and implement approaches that could potentially provide a better outcome. In our case, we could propose the following question that wasn't answered in our solution because it couldn't have greatly contributed to our recommendation.

```{r}
# When should hotels provide special offers?
# ---
hotel_df %>%
  group_by(arrival_date_month) %>% 
  summarise(n =n()) %>%
  arrange(desc(n))
```
From the observations above, March, February and November have the least bookings and therefore the hotel should have offpeak offers to encourage more bookings during this period

Our observations:
-

How does this observation tie to our solution?

## 7. Follow up questions
During this step, you rethink and propose other ways that you can improve your solution.

a). Did we have the right data?
b). Do we need other data to answer our question?
c). Did we have the right question?


